---
title: "P2data607"
author: "xiaofei"
format: html
editor: visual
---

## **Introduction**

The task is to choose three datasets from the week's class discussion to practice transforming wide-format datasets into tidy formats suitable for downstream analysis.

The three datasets I choose are: 1)world GDP by country; 2)Japanese population density; 3)a made up datsets about takeout spending habit by generation. The following process steps will be performed on each of the datasets individually:

-   Create a .CSV file that includes all of the information included in the dataset.

-   Read the information from your .CSV file into R, and use tidyr and dplyr as needed to tidy and transform your data.

-   Perform the analysis requested.

-   Code R Markdown file, posted to rpubs.com, and should include narrative descriptions of your data cleanup work, analysis, and conclusions.

## **Dataset 1: World GDP by Country: 1960-2022**

Import and load dataset. Original data has dimensions of 266 rows and 65 columns. After transformation to the long format, the new datasets has only 4 columns and 16758 rows. This long format is grouped in a way that's easier for analysis, filtering and visualization. We are ready for the next step - analysis.

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)

# create df to read the CSV file from my github repository
gdp_data <- read.csv("https://raw.githubusercontent.com/xiaofeimei1/sale-trend-analysis/refs/heads/main/GDP.csv", 
                     stringsAsFactors = FALSE,
                     check.names = FALSE)   

print("origional columns:")
print(names(gdp_data))

print("Structure of original data:")
str(gdp_data)

# Show first few rows
print("First few rows of original data:")
print(head(gdp_data))

# Transform from wide to long format
gdp_long <- gdp_data |>
  pivot_longer(
    cols = -c(Country, `Country Code`),  
    names_to = "Year",
    values_to = "GDP"
  ) |>
  mutate(
    Year = as.integer(Year),   
    GDP = as.numeric(GDP)      
  ) |>
  arrange(Country, Year)

# Show new dataset
print("Transformed data column names:")
print(names(gdp_long))

print("Structure of transformed data:")
str(gdp_long)

print("First 20 rows of transformed dataset:")
print(head(gdp_long, 20))

print("Summary of Year column:")
print(summary(gdp_long$Year))

```

Analysis and visualization. From the results shown below, The GDP dataset contains observations from 1960 to 2022. Since we GDP data listed for each year, we can do analysis on the GDP trend, such as find out which region has the top 10 growing GDP. I will also plot contry GDP but group them in regions to find out a a macro trend globally. In the end show plots to conclude analysis for this first dataset.

```{r}

# 1. BASIC SUMMARY STATISTICS
print(paste("Total countries:", length(unique(gdp_long$Country))))
print(paste("Time period:", min(gdp_long$Year, na.rm = TRUE), "to", max(gdp_long$Year, na.rm = TRUE)))
print(paste("Total observations:", nrow(gdp_long)))

# top 10 economies in the most recent year (2022)
top_economies_2022 <- gdp_long |>
  filter(Year == 2022 & !is.na(GDP)) |>
  arrange(desc(GDP)) |>
  head(10) |>
  pull(Country)

print("Top 10 economies in 2022:")
print(top_economies_2022)

# VISUALIZATIONS

# Visualization 1: Top 10 Economies Growth Over Time (1960-2022)
top_gdp_data <- gdp_long |>
  filter(Country %in% top_economies_2022 & !is.na(GDP))

ggplot(top_gdp_data, aes(x = Year, y = GDP/1e9, color = Country)) +
  geom_line(size = 1) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "GDP Growth of Top 10 Economies (1960-2022)",
       subtitle = "Log scale - Values in Billions USD",
       x = "Year", 
       y = "GDP (Billions USD, log scale)",
       color = "Country") +
  theme_minimal() +
  theme(legend.position = "bottom",
        plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  scale_x_continuous(breaks = seq(1960, 2020, by = 10))

# Global GDP Trend
global_trend <- gdp_long |>
  group_by(Year)|>
  summarise(World_GDP = sum(GDP, na.rm = TRUE) / 1e9)  # in unit billions

ggplot(global_trend, aes(x = Year, y = World_GDP)) +
  geom_line(color = "darkblue", size = 1.2) +
  geom_point(color = "darkblue", size = 0.5) +
  labs(title = "Global GDP Trend (1960-2022)",
       subtitle = "Total World GDP in Billions USD",
       x = "Year",
       y = "World GDP (Billions USD)") +
  theme_minimal() +
  scale_x_continuous(breaks = seq(1960, 2020, by = 5)) +
  scale_y_continuous(labels = scales::comma)


# Create a simple region classification and compare trend
major_economies_list <- c("United States", "China", "Japan", "Germany", "United Kingdom","France", "India", "Brazil", "Canada", "Italy", "Russia")

# Filter the data for these major economies
major_economies <- gdp_long |>
  filter(Country %in% major_economies_list & !is.na(GDP))

# Create region classification
major_economies <- major_economies |>
  mutate(Region = case_when(
    Country %in% c("United States", "Canada") ~ "North America",
    Country %in% c("Germany", "United Kingdom", "France", "Italy") ~ "Europe",
    Country == "China" ~ "China",
    Country == "India" ~ "India",
    Country == "Japan" ~ "Japan",
    Country == "Brazil" ~ "Brazil",
    Country == "Russia" ~ "Russia",
    TRUE ~ "Other"
  ))

# Plot regional comparison
ggplot(major_economies, aes(x = Year, y = GDP/1e9, color = Region, group = Country)) +
  geom_line(size = 1) +
  facet_wrap(~Region, scales = "free_y") +
  labs(title = "GDP Trends by Major Economy",
       subtitle = "Selected major economies grouped by region",
       x = "Year",
       y = "GDP (Billions USD)") +
  theme_minimal() +
  theme(legend.position = "none") +
  scale_x_continuous(breaks = seq(1960, 2020, by = 20))
```

## **Dataset 2: Japanese 2015 Population Density**

This is a dataset about Japanese population in different cities. City names were spread out in multiple columns, total 47 in the original dataset. Goal is transform dataset to long format, new column city_name will be created to group those cities in one single column, population density will be grouped under column name population_density. We will check and compare columns between original dataset with the transformed dataset in the following code chunk:

```{r}
library(tidyr)
library(dplyr)

# Read the  population density CSV file from my github repository
population_data <- read.csv("https://raw.githubusercontent.com/xiaofeimei1/sale-trend-analysis/refs/heads/main/prefecturepopdensity2015.csv", 
                            stringsAsFactors = FALSE,
                            check.names = FALSE,
                            header = TRUE)  

# original column names
print("Original columns in population dataset:")
print(names(population_data))

# structure 
print("Structure of original population data:")
str(population_data)

# Show first few rows of original data
print(head(population_data))

# Transform from wide to long format
population_long <- population_data |>
  pivot_longer(
    cols = everything(),   
    names_to = "city_name",
    values_to = "population_density"
  ) |>
  mutate(
    population_density = as.numeric(population_density),  
    city_name = as.character(city_name)   
  ) |>
  arrange(city_name)   

# Show transformed dataset results
print("Transformed data column names:")
print(names(population_long))

print("Structure of transformed data:")
str(population_long)

print("First 20 rows of transformed dataset:")
print(head(population_long, 20))

print(paste("Total rows:", nrow(population_long)))
print(paste("Number of prefectures:", length(unique(population_long$city_name))))

```

After getting the long format, we are ready to do population analysis and visualization. As shown below, top cities that has most dense population are: Tokyo, Osaka, Kanagawa... On the other hand, lowest cities with population density are Hokkaido, Iwate, Akita... To show how population were largely concentrate on the top 3 cities, we can see the last visualization that those 3 cities are far more dense compare to other cities.

```{r}

top_dense <- population_long |>
  arrange(desc(population_density)) |>
  head(10)

bottom_dense <- population_long |>
  arrange(population_density) |>
  head(10)

# Bar chart of top 10 most dense prefectures
ggplot(top_dense, aes(x = reorder(city_name, population_density), y = population_density)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Most Densely Populated Japanese Prefectures",
       subtitle = "Population Density (persons/km²) - 2015",
       x = "Prefecture",
       y = "Population Density (persons/km²)") +
  theme_minimal() +
  geom_text(aes(label = round(population_density, 0)), hjust = -0.1, size = 3)

# Bar chart of bottom 10 least dense prefectures
ggplot(bottom_dense, aes(x = reorder(city_name, desc(population_density)), y = population_density)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  labs(title = "Top 10 Least Densely Populated Japanese Prefectures",
       subtitle = "Population Density (persons/km²) - 2015",
       x = "Prefecture",
       y = "Population Density (persons/km²)") +
  theme_minimal() +
  geom_text(aes(label = round(population_density, 0)), hjust = -0.1, size = 3)

# Dot plot of all prefectures
ggplot(population_long, aes(x = reorder(city_name, population_density), y = population_density)) +
  geom_point(size = 3, color = "darkblue") +
  coord_flip() +
  labs(title = "Population Density of All Japanese Prefectures",
       subtitle = "Ordered from lowest to highest density",
       x = "Prefecture",
       y = "Population Density (persons/km²)") +
  theme_minimal() +
  geom_hline(yintercept = mean(population_long$population_density), 
             linetype = "dashed", color = "red", size = 0.5) +
  annotate("text", x = 5, y = mean(population_long$population_density) + 100, 
           label = "National Average", color = "red", size = 3)



```

## **Dataset 3:** takeout spending habit by generation.

The original takeout spending dataset contained 12 rows and 7 columns, with 5 generations: Gen Z, Millennials, Gen X, Baby Boomers, Silent Gen. and their monthly spending columns from January to June. After transformation, the dataset was reshaped into long format with 30 rows and 3 columns: generation, month and spending. This transformed datasets is in tidy structure are are ready for data analysis and visualization.

```{r}

# Read the takeout spending CSV file
takeout_data <- read.csv("https://raw.githubusercontent.com/xiaofeimei1/sale-trend-analysis/refs/heads/main/takeoutspending.csv", 
                         stringsAsFactors = FALSE,
                         check.names = FALSE)

# Print original columns 
print(names(takeout_data))

# Show structure of original data
str(takeout_data)

# Show first few rows of original data
print("First few rows of original takeout data:")
print(head(takeout_data))

# Clean the data first - remove empty rows
takeout_data_clean <- takeout_data |>
  filter(!is.na(generation) & generation != "") |>  # Remove rows with empty generation
  select(1:7)  

# Transform from wide to long 
takeout_long <- takeout_data_clean |>
  pivot_longer(
    cols = -generation,  # All month columns except generation
    names_to = "month",
    values_to = "spending"
  ) |>
  mutate(
    month = as.character(month),  #  month in character
    spending = as.numeric(spending),  #  spending is numericr
    month = factor(month, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun"), ordered = TRUE)
  ) |>
  arrange(generation, month)  
# Show transformed dataset results
print(names(takeout_long))

print("Structure of transformed data:")
str(takeout_long)

print("All rows of transformed dataset:")
print(takeout_long)


print(paste("Total rows:", nrow(takeout_long)))
print(paste("Number of generations:", length(unique(takeout_long$generation))))
print(paste("Months covered:", paste(levels(takeout_long$month), collapse = ", ")))

# average spending by generation
print("Average spending by generation:")
takeout_long |>
  group_by(generation) |>
  summarise(avg_spending = mean(spending, na.rm = TRUE)) %>%
  arrange(desc(avg_spending))

# average spending by month
print("Average spending by month:")
takeout_long |>
  group_by(month) |>
  summarise(avg_spending = mean(spending, na.rm = TRUE))
```

Data analysis and visualization to identify which generation spends the most on takeout per month and to find out monthly spending trends across different generational groups.

The results shows that Millennials lead in takeout spending, with the highest average monthly expenditures. They are closely followed by Gen Z and Gen X, whose spending levels are very similar to each other.

Regarding monthly trends, the analysis shows a upward pattern from January to June across all generations, indicating that takeout spending generally increases as the year progresses from winter to summer months. This seasonal trend, combined with generational differences, could be valuable infomation to study consumer behavior and for company to potential targeting specific generation to promote their food products.

```{r}
generation_summary <- takeout_long |>
  group_by(generation) |>
  summarise(
    Avg_Monthly = mean(spending, na.rm = TRUE),
    .groups = 'drop'
  ) |>
  arrange(desc(Avg_Monthly))

# Bar chart - Average monthly spending by generation
ggplot(generation_summary, aes(x = reorder(generation, Avg_Monthly), y = Avg_Monthly, fill = generation)) +
  geom_bar(stat = "identity", width = 0.7) +
  coord_flip() +
  labs(title = "Average Monthly Takeout Spending by Generation",
       subtitle = "Based on Jan-Jun 2015 data",
       x = "Generation",
       y = "Average Monthly Spending ($)") +
  theme_minimal() +
  theme(legend.position = "none") +
  geom_text(aes(label = paste0("$", round(Avg_Monthly, 0))), hjust = -0.2, size = 4) +
  scale_fill_brewer(palette = "Set2") +
  ylim(0, max(generation_summary$Avg_Monthly) * 1.1)

# Line chart - Spending trends over months by generation
ggplot(takeout_long, aes(x = month, y = spending, group = generation, color = generation)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Takeout Spending Trends by Generation (Jan-Jun 2015)",
       subtitle = "Monthly spending patterns across generations",
       x = "Month",
       y = "Spending ($)",
       color = "Generation") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_color_brewer(palette = "Set1")

#Comparison across months
ggplot(takeout_long, aes(x = month, y = spending, fill = generation)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.7) +
  labs(title = "Monthly Takeout Spending by Generation",
       subtitle = "Side-by-side comparison",
       x = "Month",
       y = "Spending ($)",
       fill = "Generation") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  scale_fill_brewer(palette = "Set2")

```

## **Summary**

This data analysis project focused on transforming three distinct datasets (global GDP, Japanese prefecture population density, and generational takeout spending) from wide format to long format to enable more effective analysis and visualization. Wide format has data of same category across multiple columns, making comparative analysis more difficult. The transformed data made filtering easier, enhanced grouping and aggregation when doing analysis. This data transformation to get tidy data is essential part for complex data analysis.
